Optimal Fitting of Planar Curves to Prescribed Constraints

### 2014-04-02 idp/documents/documentation/paper.tex
added some corrections to idp paper

		The scope of this project is restricted to two-dimensional vector graphics, but many ideas can be applied to three-dimensional vector graphics as well. All elements in two-dimensional vector graphics are described in terms of curves, so the design process of curves is the single most important part of two-dimensional vector graphics. Unfortunately, designing curves using the tools available in current vector graphics software can be very frustrating. The user is often unable to make the curve look the way they want it to. To make matters worse it is often unclear why this is the case, as the software seems to put all the necessary tools into the hands of the designer, yet at times, they somehow do not allow the user to communicate their intention to the software.

		While there has been some promising research into improving curve design tools \cite{thesis-mvc} \cite{thesis-spiro}, it focuses mostly on the aspect of improving the fairness of curves, rather than usability as a whole. Unfortunately, it also has not had much of an impact on actual vector graphics software used in practice, which continues to focus almost exclusively on Bézier splines to this day.

		Thus, the objective of this project is to identify the exact nature and cause of the shortcomings of current curve design tools in order to use this knowledge to develop a curve design tool that does not exhibit them.

@@

		When trying to improve something, it seems advisable to take a step back and thoroughly research the causes of its perceived shortcomings so as to find a way to improve it without making the same or similar mistakes. In the context of curve design tools, this can be achieved by analyzing and modeling the process that people use when designing curves. Once such a model has been obtained, it can be used to propose criteria for good curve design tools as well as to analyze usability aspects of existing tools.

@@

			Step 1 is of course independent of any software that may be used by the designer. Step 2 and 3 are strongly influenced by the choice of design tool, in that the software dictates which properties it uses for the specification of curves (the specification language the software uses). Unless the properties the designer conveyed to the software in step 3 uniquely identify the curve, in step 4, the design tool will have to choose the curve that the user was most likely describing from a possibly infinite set of curves that exhibit the supplied properties. This notion of `most likely' is usually modeled in terms of a fairness measure, fairer curves being regarded as more likely. Fairness usually encodes intuitive notions such as smoothness and minimality using mathematical concepts such as continuity and variation of curvature. Here, the software has another opportunity to influence the design process both positively and negatively depending on what fairness measure is chosen and how good the software is at reliably finding a curve that has a high fairness.

			Note that there may be multiple ways to describe a given curve design tool in terms of this process. Instead of modeling the selection of the most likely curve using some fairness measure, one may also propose an alternate view on the process in which the user implicitly specifies the exact curve they want in the properties they specify. For example, in a curve design tool which takes two points from the user and connects them using a straight line, one may say that the user provides the start and the end point of the curve, and the fairness measure is designed to always select the straight line as the fairest curve. Another way of looking at this process is to say that the user, by supplying \(p_0\) and \(p_1\), specified the coefficients of the linear term \(p_0 + \xp{p_1 - p_0} t\), thus uniquely identifying the desired curve. Modeling the process in such a way that the user supplies only the minimum amount of implicit information, relying on the fairness measure for everything else, is usually closer to reality when analyzing usability of some curve design tool.

			In a similar fashion, one may consider some shortcomings (for instance, insufficient continuity) as either shortcomings of the specification language (the language does not allow the user to request curvature continuity) or as shortcomings of the fairness measure (the fairness measure does not select sufficiently smooth curves). These decisions have to be made on a case by case basis, selecting the approach that is better suited to modeling the reality of curve design.

@@

				Applying the criteria established in section \ref{section:usability_criteria_curve_design_tools}, we conclude that the design tools for Bézier splines work sufficiently well in terms of humans trying to read and/or speak their specification language (criteria 2 and 3), a skill that can be acquired with some practice. The curve chosen by the fairness measure is also reasonably minimal (criterion 5), since the spline's complexity is limited by the number of segments that are used, which is in direct correspondence with the specification. Bézier splines are furthermore very simple to derive from their description, fulfilling criterion 6 perfectly. We think that design tools for Bézier splines have some issues with both the fairness measure's smoothness (criterion 4) and the expressiveness of the description language (criterion 1), both of which shall be analyzed in the following paragraphs.

				The issue with the fairness measure is that it always selects the uniquely determined cubic Bézier curve for each segment, thus not guaranteeing curvature continuity across segments. It is difficult for humans to choose the coefficient points in such a way that curvature continuity is established, making it hard to create smooth curves if the design software does not support the user in this regard. The problem is aggravated by the fact that curvature discontinuities can be hard to spot and may thus only be discovered much later in the design process, increasing the cost of these mistakes. In case the software allows the user to request curvature continuous Bézier splines, this sacrifices some degree of freedom for each node, thus restricting the way nodes can be manipulated and in turn possibly forcing the user to place more nodes in order to get the desired result. Placing more nodes leads to some disadvantages that will be discussed further in the paragraphs dealing with description language expressiveness.

				The description language used to specify Bézier splines in common design tools is lacking in expressiveness. Specifically, describing curves in terms of their curvature is difficult since curvature may only be specified indirectly through the magnitude of the velocity values (higher velocity resulting in lower curvature). Unfortunately, this is also very dependent on the point and velocity coefficients at the other end of the spline segment. There is also no way to directly specify constant or linearly changing curvature. This makes specifying curvature very tedious, and in case the winding angle of the curve segment is large enough, actually impossible, forcing the designer to use additional nodes in order to get the desired result. Using an excessive number of nodes makes it more tedious and time-consuming for the designer to create and/or modify such curves, since many nodes and handles have to be created or moved in order to change the overall curve. It also sacrifices smoothness, since it is easy to accidentally introduce subtle bumps in the curve when using a large number of nodes. Another issue is that without software support for curvature continuity, using more nodes usually also introduces more points where curvature is not continuous. In cases where the curvature is an important property of the curve (for example the constant curvature of circular arcs, or the linearly changing curvature of spirals), it may be very difficult to specify these properties indirectly using the coefficient points.

@@

				Figure \ref{figure:examples_bézier_arc} shows two Bézier splines. The left curve is a Bézier spline approximation of a circular arc, exhibiting nearly constant curvature, while the right one is not, going from low curvature to high curvature and back to low curvature. Since the difference can be hard to see, it is easy to make mistakes when trying to approximate circular arcs using Bézier splines.

@@

				The red curve segment in figure \ref{figure:examples_bézier_lugia} is a Bézier spline with a curvature discontinuity which went unnoticed by the artist at the time of creation. In fact, this discontinuity would be even less visible if the curve was not vertical at said point, manifesting itself only in a reduced overall smoothness of the curve, without a clear indication of where the problem lies.

@@

				Spiro splines were added to Inkscape in 2009. Even though they were integrated using path effects and thus have a somewhat cumbersome user interface, they constitute a commendable effort of incorporating results from curve research into vector graphics software, something that has hardly happened this far. More information on Spiro splines can be found in the dissertation of Raphael Levien \cite{thesis-spiro}.

@@

				Again, the model from section \ref{section:curve_design_process} is applied. With Spiro splines, the modeling is fairly straightforward. The user supplies a set of points for which the software computes an interpolating spline. The fairness measure chooses the segments of this spline to be parts of the Euler spiral and makes sure that there is curvature continuity and no excessive winding.

				We analyze Spiro splines in terms of the criteria presented in section \ref{section:usability_criteria_curve_design_tools}. The specification language of Spiro splines being simply points on the curve, plus the fact that on sufficiently smooth curves, not very many of them are needed, the language is well-suited to be read and spoken by humans (criteria 2 and 3). The fairness measure leads to beautifully smooth curves that are usually sufficiently close to what a human would expect given the control points (criteria 4 and 5). The software is also good and sufficiently quick at constructing these curves (there are some cases where the algorithm fails to find a meaningful curve, but these cases are fairly artificial and can usually be avoided when designing actual curves) and thus fulfills criterion 6 well enough. The only real issue lies in the insufficient expressiveness of the description language (criterion 1), discussed further in the following paragraph.

				Section \ref{section:bézier_splines} discussed how the lack of ability to specify curvature using common Bézier spline design tools adversely affects usability. Many of the same points apply to Spiro spline design tools, since they allow the user to neither specify direction nor curvature. These properties can only be expressed indirectly, by placing the control points in such a way that the algorithm will create a curve that has the desired properties, resulting in the usability issues described previously, such as requiring many control points to be specified even though the curve could be described much more concisely in terms of direction and/or curvature. This seems especially unfortunate when considering that the spline primitive of Spiro splines, the Euler spiral, would lend itself excellently to the expression of circular arcs, spirals and other shapes exhibiting constant or linearly changing curvature. Sometimes, the algorithm can be coaxed into producing the desired curve by placing just a few control points very carefully, resulting in a series of Euler spiral segments that exhibit the wanted properties. Unfortunately, this is rarely practical, since it both requires a lot of time to set up and does not behave very well when modified. These issues are somewhat alleviated by the guaranteed curvature continuity together with the general tendency of the algorithm to produce extremely smooth curves even when many control points are involved. That way, when indirectly specifying tangent angle and/or curvature by placing many control points, it is not as easy to accidentally create small bumps as with Bézier splines.

@@

				Figure \ref{figure:examples_spiro_fluttershy} shows a shape that is in principle well-suited for being represented by parts of the Euler spiral. Unfortunately, describing the curve using Spiro splines still requires many control points. Note that even though this example has the same number of specified on-curve points (7) as the corresponding example in section \ref{section:bézier_splines}, the latter has an additional 12 velocity specifications, proving that even though curvature cannot be specified explicitly, Spiro splines are a much better choice for shapes with approximately linearly changing curvature than Bézier splines.

@@

		Having analyzed the usability aspect of curve design tools in section \ref{section:problem_analysis}, it is time to think about how to properly address their current issues. In the following sections, we propose both an abstract approach for designing usability-oriented curve design tools, as well as specific ideas for developing a piece of software using this approach.

@@

			We think that in order to get the best usability, using a good specification language which does not rely on implementation details has to be the top priority. Together with a good fairness measure, it forms a description language that, if it allows for reliable and efficient derivation of curves, will then lead to usable curve design tools. Since it is not immediately clear which description language is the best, the first step is to move away from the traditional explicit curve construction, and instead use a process that enables the creation of curves based on descriptions.

@@

				The specification language we propose uses differential geometric properties of parametric curves and is thus decoupled from specifics of the underlying curve model, hopefully leading to a specification language that is easier to learn and use. The specification language will focus on the properties point, direction and curvature as defined in \ref{section:parametric_plane_curves}. In order to provide maximal flexibility, the specification language should allow to specify any combination of these three properties at any number of positions on the curve.

				When only concerned with point specifications, it may be enough to require the curve to go through all of them in order, although sometimes, even this will lead to undesirable results. That is because even with the curve preserving the order in which the points are visited, it is unspecified how much arc length the curve will spend between each pair of points. When specifying direction and/or curvature without an associated point, it becomes a necessity to associate these specifications with some kind of positional information in order to anchor them to points on the curve. If this was not done, the specifications could move freely along the curve and would thus behave very unpredictably. We thus chose to associate each combination of point, direction and curvature with a position on the curve.

				With that in mind, it is now possible to abandon this notion of a combination of point, direction and curvature specifications and instead talk about specification items, a specification item being a positioned point, direction or curvature specification. For example, a combination of point and curvature specification at some position thus splits up into a point specification item at this position and a curvature specification item at the same position.

				An obvious choice for specifying a position on a curve would be by giving a parameter value for the underlying parametric curve. However, this would violate the design idea of using differential geometric properties of curves. A way of specifying a position on a curve that is independent from its parametrization is the arc length of the curve. Instead of directly using the arc length to give the positions of specification items, we chose to specify positions as fractions of the total arc length of the curve. This way, the position of a specification item relative to the start and end points of the curve is independent of the total arc length of the curve.

				With this setup, it seems advisable to also include the curve's total arc length in the specification. This way, one can sensibly specify curves that have only one point specification item. Also, with unspecified curve length, changing any specification item may inadvertently change the curve's total arc length, thus possibly adversely affecting other parts of the curve whose specification item positions are given as a fraction of the curve's total arc length. For instance, with three point specification items on the curve, pulling the third one further away from the second one would increase the total arc length of the curve, thus also increasing the arc length of the curve segment between the first and the second point specification item, thereby unintentionally creating a bulge between those two specification items.

				Thus, the final specification language is as follows:
				\begin{equation*}
					\begin{alignedat}{2}
						& \mathrm{Positions}          && = \unit\\
						& \mathrm{Points}             && = \xc{1} \times \R{2}\\
						& \mathrm{Directions}         && = \xc{2} \times \RR\\
						& \mathrm{Curvatures}         && = \xc{3} \times \RR\\
						& \mathrm{SpecificationItems} && = \mathrm{Positions} \times \left(\mathrm{Points} \cup \mathrm{Directions} \cup \mathrm{Curvatures}\right)\\
						& \mathrm{CurveLengths}       && = \RR\\
						& \mathrm{Specifications}     && = \powerset{\mathrm{SpecificationItems}} \times \mathrm{CurveLengths}
					\end{alignedat}
				\end{equation*}

@@

				In order to address the issues of the MEC, there has also been some research into fairness measures whose design is more directly motivated by the human perception of smoothness in plane curves. One such approach is that of the minimum variation curve (MVC) \cite{thesis-mvc}, which tries to minimize the variation of the curvature. The corresponding functional is given below:
				\begin{equation*}
					\integral{0}{1}{\apply{\chi'}{t}^2}{t}
				\end{equation*}

				Minimum variation curves capture the intuitive notion of smoothness very well. The functional will put straight lines through points that are co-linear and circular arcs through points that are co-circular when otherwise unconstrained. When constraints cause different points on the curve to have different curvatures, the fairness measure will prefer linearly changing curvature, since that will make the derivative of the curvature a constant and thereby minimize the integral over its square. In general, the MVC fairness measure is very good at avoiding bumps and inflection points when possible. It also works nicely with curvature specifications, since constant curvature is perfectly acceptable for the MVC fairness measure.

@@

				The implementation we chose uses smooth optimization, or, more specifically, an interior point method. This type of algorithm further requires the functions \(f\) and \(g\) to be at least once, ideally twice continuously differentiable.

@@

				Let \(n\) be the number of specification items. Let \(\hat{t}_i\) be the position of the \(i\)th specification item. If the \(i\)th item is a point specification item, the point is denoted \(\hat{\phi}_i\), if it is a direction specification item, the direction is denoted \(\hat{\delta}_i\) and if it is a curvature specification item, the curvature is denoted \(\hat{\chi}_i\). The specified total arc length for the curve is denoted \(\hat{\lambda}\).

@@

				Since the position of a specification item is given as a fraction of the curve's total arc length, the function mapping the covered arc length to the corresponding parameter values is needed in order to state the optimization goals for the specification items. Unfortunately, this function is not very well-suited to be used as a term in a smooth optimization problem even for simple curves like polynomials. Also, since the function is not fixed and varies with the coefficients, there is no way of telling on which segment each specification item will end up, something that is also difficult to state in a smooth optimization problem. The specification of the total arc length of the curve also needs to be stated as an optimization goal, which may be difficult if the the covered arc length function is hard to express. Furthermore, the difficulty of expressing the fairness measure, which is often an integral over the arc length of the curve, is dependent on the complexity of the covered arc length function. In order to address all these problems, the following error term is used:
				\begin{equation*}
					\max_{t \in \unit} \xa{\apply{\sigma}{t} - \hat{\lambda}}
				\end{equation*}
				For polynomial splines with sufficient degree and segment count, this term will be brought close to zero in the optimization, making the speed approximately constant with value \(\hat{\lambda}\). This, in turn, makes the covered arc length function approximately linear (\(\apply{\lambda}{t} \approx \hat{\lambda}t\)), facilitating the simple statement of the optimization goals for the specification items and the fairness measure as well as fixing the segment for each specification item. It also causes the total arc length of the curve to be approximately equal to \(\hat{\lambda}\), as requested by the specification.

@@

				The error term given above is not well-suited for smooth optimization because it incorporates finding a maximum. Since the optimization problem will allow the original error term to be brought close to zero, it can be approximated using the following error term:
				\begin{equation*}
					\hat{\lambda}^{-2}\integral{0}{1}{\xp{\apply{\sigma}{t} - \hat{\lambda}}^2}{t}
				\end{equation*}
				Since this term is intended to be used as a minimization goal, the factor \(\hat{\lambda}^{-2}\) was added, which makes it scale invariant. This enables the error term to be statically weighted against other terms in the objective function without this weighting being shifted because of curve scaling.

@@

				As mentioned in section \ref{section:description_language}, we are not concerned with properties beyond the second derivative of the curve. Thus, only \(\contg{2}\) continuity will be required. While the polynomial curves used as segments are smooth (\(\contp{\infty}\) continuity), the resulting spline in general is not. The segments have to be pieced together in such a way as to ensure a certain degree of smoothness. Thus, to ensure \(\contg{2}\) continuity for the whole curve, it is sufficient to ensure \(\contg{2}\) continuity at the segment connection points. To do this, the following error terms are added for each connection point (\(0 \leq i \leq m - 1\)):
				\begin{equation*}
					\begin{gathered}
						\apply{\phi_i}{1} - \apply{\phi_{i + 1}}{0}\\
						\apply{\phi'_i}{1} - \apply{\phi'_{i + 1}}{0}\\
						\apply{\phi''_i}{1} - \apply{\phi''_{i + 1}}{0}
					\end{gathered}
				\end{equation*}
				These error terms being zero implies \(\contp{2}\) continuity. However, since the constant speed requirement introduced in the previous paragraph collapses \(\contp{2}\) continuity and \(\contg{2}\) continuity, this is equivalent to the original requirement of ensuring only \(\contg{2}\) continuity, yet better suited for optimization.

				Since it is not acceptable for the curve segments to not touch, have sharp corners or curvature discontinuities, the error terms above are added as zero-error constraints.

@@

				Having gathered all the necessary error terms, it is time to put them together. Terms with zero-error constraints, as well as terms with upper bounds on the error are added as component functions of the constraint function \(g\) together with appropriate bound components in \(g_l\) and \(g_u\).

				Minimization goals, on the other hand, are assembled in a weighted sum which defines the the objective function \(f\). Since the error terms were designed to be scale invariant, this weighting can be done statically. The exact values of the weights are to be determined empirically.

>>>>

		The scope of this project is restricted to two-dimensional vector graphics, but many ideas can be applied to three-dimensional vector graphics as well. All elements in two-dimensional vector graphics are described in terms of curves, so the design process of curves is the single most important part of two-dimensional vector graphics. Unfortunately, designing curves using the tools available in current vector graphics software can be very frustrating, with the user being unable to achieve the desired curve shape. To make matters worse it is often unclear why this is the case, as the software seems to put all the necessary tools into the hands of the designer, yet at times, they somehow do not allow the user to communicate their intention to the software.

		While there has been some promising research into improving curve design tools \cite{thesis-mvc} \cite{thesis-spiro}, it focuses mostly on the aspect of improving the fairness of curves, rather than usability as a whole. Additionally, it has not had much of an impact on actual vector graphics software used in practice, which continues to focus almost exclusively on Bézier splines to this day.

		Thus, the objective of this project is to identify the exact nature and cause of the shortcomings of current curve design tools in order to use this knowledge to develop better ones.

@@

		When trying to improve something, it seems advisable to take a step back and thoroughly research the causes of its perceived shortcomings in order to find a way to improve it without making the same or similar mistakes. In the context of curve design tools, this can be achieved by analyzing and modeling the process that people use when designing curves. Once such a model has been obtained, it can be used to propose criteria for good curve design tools as well as to analyze usability aspects of existing tools.

@@

			Step 1 is of course independent of any software that may be used by the designer. Step 2 and 3 are strongly influenced by the choice of design tool, as the software dictates which properties it uses for the specification of curves (the specification language the software uses). Unless the properties the designer conveyed to the software in step 3 uniquely identify the curve, in step 4, the design tool will have to choose the curve that the user was most likely describing from a possibly infinite set of curves that exhibit the supplied properties. This notion of `most likely' is usually modeled in terms of a fairness measure, fairer curves being regarded as more likely. Fairness usually encodes intuitive notions such as smoothness and minimality using mathematical concepts such as continuity and variation of curvature. At this point, the software has another opportunity to influence the design process both positively and negatively depending on what fairness measure is chosen as well as the ability of the software to find a curve that has a high fairness.

			Note that there may be multiple ways to describe a given curve design tool in terms of this process. Instead of modeling the selection of the most likely curve using some fairness measure, one may also propose an alternate view on the process in which the user implicitly specifies the exact curve they want in the properties they specify. For example, in a curve design tool which takes two points from the user and connects them using a straight line, one possible view is that the user provides the start and the end point of the curve, and the fairness measure is designed to always select the straight line as the fairest curve. Another way of looking at this process is to say that the user, by supplying \(p_0\) and \(p_1\), specified the coefficients of the linear term \(p_0 + \xp{p_1 - p_0} t\), thus uniquely identifying the desired curve. Modeling the process in such a way that the user supplies only the minimum amount of implicit information, relying on the fairness measure for everything else, is usually closer to reality when analyzing usability of some curve design tool.

			In a similar fashion, one may attribute certain shortcomings (for instance, insufficient continuity) to either the specification language (the language does not allow the user to request curvature continuity) or the fairness measure (the fairness measure does not select sufficiently smooth curves). These decisions have to be made on a case by case basis, selecting the approach that is better suited to modeling the reality of curve design.

@@

				Applying the criteria established in section \ref{section:usability_criteria_curve_design_tools}, we conclude that the design tools for Bézier splines work sufficiently well in terms of humans trying to read and/or speak their specification language (criteria 2 and 3), a skill that can be acquired with some practice. The curve chosen by the fairness measure is reasonably minimal (criterion 5), as the spline's complexity is limited by the number of segments that are used, which is in direct correspondence with the specification. Bézier splines are furthermore very simple to derive from their description, fulfilling criterion 6 perfectly. We think that design tools for Bézier splines have some issues with both the fairness measure's smoothness (criterion 4) and the expressiveness of the description language (criterion 1), both of which shall be analyzed in the following paragraphs.

				The issue with the fairness measure is that it always selects the uniquely determined cubic Bézier curve for each segment, thus not guaranteeing curvature continuity across segments. It is difficult for humans to choose the coefficient points in such a way that curvature continuity is established, making it hard to create smooth curves if the design software does not support the user in this regard. The problem is aggravated by the fact that curvature discontinuities can be hard to spot and may thus only be discovered much later in the design process, increasing the cost of these mistakes. In case the software does support the user by allowing them to request curvature continuous Bézier splines, this sacrifices some degree of freedom for each node, thus restricting the way nodes can be manipulated and in turn possibly forcing the user to place more nodes in order to get the desired result. Placing more nodes leads to some disadvantages that will be discussed further in the next paragraph.

				The description language used to specify Bézier splines in common design tools lacks expressiveness. Specifically, describing curves in terms of their curvature is difficult since curvature may only be specified indirectly through the magnitude of the velocity values (higher velocity resulting in lower curvature). Unfortunately, this is also very dependent on the point and velocity coefficients at the other end of the spline segment. Additionally, there is no way to directly specify constant or linearly changing curvature. This makes specifying curvature very tedious, and in case the winding angle of the curve segment is large enough, actually impossible, forcing the designer to use additional nodes in order to achieve the desired result. Using an excessive number of nodes makes it more tedious and time-consuming for the designer to create and/or modify such curves, as many nodes and handles have to be created or moved in order to change the overall curve. It also sacrifices smoothness, since it is easy to accidentally introduce subtle bumps in the curve when using a large number of nodes. Another issue is that without software support for curvature continuity, using more nodes usually also introduces more points where curvature is not continuous. In cases where the curvature is an important property of the curve (for example the constant curvature of circular arcs, or the linearly changing curvature of spirals), it may be very difficult to specify these properties indirectly using the coefficient points.

@@

				Figure \ref{figure:examples_bézier_arc} shows two Bézier splines. The left curve is a Bézier spline approximation of a circular arc, exhibiting nearly constant curvature, while the right one is not, changing from low curvature to high curvature and back to low curvature. As the difference can be hard to see, it is easy to make mistakes when trying to approximate circular arcs using Bézier splines.

@@

				The red curve segment in figure \ref{figure:examples_bézier_lugia} is a Bézier spline with a curvature discontinuity which went unnoticed by the artist at the time of creation. In fact, this discontinuity would be even less visible if the curve was not vertical at said point, manifesting itself only in a reduced overall smoothness of the curve, without a clear indication of the problem's origin.

@@

				Spiro splines \cite{thesis-spiro} were added to Inkscape in 2009. Even though they were integrated using path effects and thus have a somewhat cumbersome user interface, they constitute a commendable effort of incorporating results from curve research into vector graphics software, something that has hardly happened this far.

@@

				In order to evaluate the usability of Spiro splines, the model from section \ref{section:curve_design_process} is applied once more. In the case of Spiro splines, the modeling is fairly straightforward. The user supplies a set of points for which the software computes an interpolating spline. The fairness measure chooses the segments of this spline to be parts of the Euler spiral, ensuring curvature continuity and trying to avoid excessive winding.

				We analyze Spiro splines in terms of the criteria presented in section \ref{section:usability_criteria_curve_design_tools}. With the specification language of Spiro splines consisting of points on the curve, plus the fact that on sufficiently smooth curves, not very many of them are needed, the language is well-suited to be read and spoken by humans (criteria 2 and 3). The fairness measure leads to beautifully smooth curves that are usually sufficiently close to what a human would expect given the control points (criteria 4 and 5). The software is also reliable and quick with the construction of these curves. There are some cases where the algorithm fails to find a meaningful curve, but these cases are fairly artificial and can usually be avoided when designing actual curves. It thus fulfills criterion 6 satisfactorily. The only real issue lies in the insufficient expressiveness of the description language (criterion 1), discussed further in the following paragraph.

				Section \ref{section:bézier_splines} discussed how the lack of ability to specify curvature using common Bézier spline design tools adversely affects usability. Many of the same points apply to Spiro spline design tools, since they allow the user to neither specify direction nor curvature. These properties can only be expressed indirectly, by placing the control points in such a way that the algorithm will create a curve that has the desired properties. This results in the usability issues described previously, such as requiring many control points to be specified even though the curve could be described much more concisely in terms of direction and/or curvature. This seems especially unfortunate, considering that the spline primitive of Spiro splines, the Euler spiral, would lend itself excellently to the expression of circular arcs, spirals and other shapes exhibiting constant or linearly changing curvature. Sometimes, the algorithm can be coaxed into producing the desired curve by placing just a few control points very carefully, resulting in a series of Euler spiral segments that exhibit the wanted properties. Unfortunately, this is rarely practical, since it both requires a lot of time to set up and does not behave very well when modified. These issues are somewhat alleviated by the guaranteed curvature continuity and the general tendency of the algorithm to produce extremely smooth curves even when many control points are involved. That way, when indirectly specifying tangent angle and/or curvature by placing many control points, it is not as easy to accidentally create small bumps as with Bézier splines.

@@

				Figure \ref{figure:examples_spiro_fluttershy} shows a shape that is in principle well-suited for being represented by parts of the Euler spiral. Unfortunately, describing the curve using Spiro splines still requires many control points. Note that even though this example has the same number of specified on-curve points (7) as the corresponding example for Bézier splines shown in figure \ref{figure:examples_bézier_fluttershy}, the latter has an additional 12 velocity specifications. This illustrates that even though curvature cannot be specified explicitly, Spiro splines are a much better choice for shapes with approximately linearly changing curvature than Bézier splines.

@@

		Having analyzed the usability aspect of curve design tools in section \ref{section:problem_analysis}, it is time to think about how to properly address their current issues. In the following sections, we propose both an abstract approach for designing usability-oriented curve design tools, as well as specific ideas for developing software using this approach.

@@

			We think that in order to get good usability, using a powerful specification language which does not rely on implementation details has to be the top priority. In combination with a good fairness measure, it forms a description language that, if it allows for reliable and efficient derivation of curves, will then lead to usable curve design tools. Since it is not immediately clear which description language is the best, the first step is to move away from the traditional explicit curve construction, and instead use a process that enables the creation of curves based on descriptions.

@@

				The specification language we propose uses differential geometric properties of parametric curves and is thus decoupled from specifics of the underlying curve model, hopefully leading to a specification language that is easier to learn and use. The specification language will focus on the properties point, direction and curvature as defined in section \ref{section:parametric_plane_curves}. In order to provide maximal flexibility, the specification language should allow to specify any combination of these three properties at any number of positions on the curve.

				When only concerned with point specifications, requiring the curve to go through all of them in order may be enough to allow the user to express their intent, although sometimes, even this will lead to undesirable results. That is because even with the curve preserving the order in which the points are visited, it is unspecified how much arc length the curve will spend between each pair of points. When specifying direction and/or curvature without an associated point, it becomes a necessity to associate these specifications with some kind of positional information in order to anchor them to points on the curve. If this were not done, the specifications could move freely along the curve and would therefore behave very unpredictably. We thus chose to associate each combination of point, direction and curvature with a position on the curve.

				With that in mind, it is now possible to abandon this notion of a combination of point, direction and curvature specifications and instead talk about specification items, a specification item being a positioned point-, direction- or curvature specification. For example, a combination of point and curvature specification at some position thus splits up into a point specification item at this position and a curvature specification item at the same position.

				The most simple way of specifying a position on a curve would be by giving a parameter value for the underlying parametric curve. However, this would violate the design idea of using differential geometric properties of curves. A way of specifying a position on a curve that is independent from its parametrization is the arc length of the curve. Instead of directly using the arc length to describe the positions of specification items, we chose to specify positions as fractions of the total arc length of the curve. This way, the position of a specification item relative to the start and end points of the curve is independent of the total arc length of the curve.

				With this setup, it seems advisable to also include the curve's total arc length in the specification. This way, one can sensibly specify curves that have only one point specification item. Also, with unspecified curve length, changing any specification item may inadvertently change the curve's total arc length, thus possibly adversely affecting other parts of the curve whose specification item positions are given as a fraction of the curve's total arc length. For instance, with three point specification items on the curve, pulling the third one further away from the second one would increase the total arc length of the curve. This would also increase the arc length of the curve segment between the first and the second point specification item, thereby unintentionally creating a bulge between those two specification items.

				One way to formally represent the final specification language is as follows:
				\begin{equation*}
					\begin{alignedat}{2}
						& \mathrm{Positions}          && = \unit\\
						& \mathrm{Points}             && = \xc{1} \times \R{2}\\
						& \mathrm{Directions}         && = \xc{2} \times \RR\\
						& \mathrm{Curvatures}         && = \xc{3} \times \RR\\
						& \mathrm{SpecificationItems} && = \mathrm{Positions} \times \left(\mathrm{Points} \cup \mathrm{Directions} \cup \mathrm{Curvatures}\right)\\
						& \mathrm{CurveLengths}       && = \RR\\
						& \mathrm{Specifications}     && = \powerset{\mathrm{SpecificationItems}} \times \mathrm{CurveLengths}
					\end{alignedat}
				\end{equation*}

@@

				In order to address the issues of the MEC, there has been some research into fairness measures whose design is motivated more directly by the human perception of smoothness in plane curves. One such approach is that of the minimum variation curve (MVC) \cite{thesis-mvc}, which tries to minimize the variation of the curvature. The corresponding functional is given below:
				\begin{equation*}
					\integral{0}{1}{\apply{\chi'}{t}^2}{t}
				\end{equation*}

				Minimum variation curves capture the intuitive notion of smoothness very well. The functional will put straight lines through points that are co-linear and circular arcs through points that are co-circular when otherwise unconstrained. When constraints cause different points on the curve to have different curvatures, the fairness measure will prefer linearly changing curvature, as this will result in the derivative of the curvature being constant, thereby minimizing the integral over its square. In general, the MVC fairness measure is very good at avoiding bumps and inflection points when possible. It also works nicely with curvature specifications, since constant curvature is perfectly acceptable for the MVC fairness measure.

@@

				The implementation we chose uses smooth optimization, specifically an interior point method. This type of algorithm further requires the functions \(f\) and \(g\) to be at least once, ideally twice continuously differentiable.

@@

				In the following sections, let \(n\) be the number of specification items. Let \(\hat{t}_i\) be the position of the \(i\)th specification item. If the \(i\)th item is a point specification item, the point is denoted \(\hat{\phi}_i\), if it is a direction specification item, the direction is denoted \(\hat{\delta}_i\) and if it is a curvature specification item, the curvature is denoted \(\hat{\chi}_i\). The specified total arc length for the curve is denoted \(\hat{\lambda}\).

@@

				As established in section \ref{section:specification_language}, the position of a specification item is given as a fraction of the curve's total arc length, the function mapping the covered arc length to the corresponding parameter values is needed in order to state the optimization goals for the specification items. Unfortunately, this function is not very well-suited to be used as a term in a smooth optimization problem even for simple curves like polynomials. Also, the function is not fixed and varies with the coefficients, so there is no way of telling on which segment each specification item will end up, something else that is difficult to state in a smooth optimization problem. Another problem is that the specification of the total arc length of the curve needs to be stated as an optimization goal, which may be difficult if the the covered arc length function is hard to express. Furthermore, the difficulty of expressing the fairness measure, which is often an integral over the arc length of the curve, is dependent on the complexity of the covered arc length function. In order to address all these problems, the following error term is used:
				\begin{equation*}
					\max_{t \in \unit} \xa{\apply{\sigma}{t} - \hat{\lambda}}
				\end{equation*}
				For polynomial splines with sufficient degree and segment count, this term will be brought close to zero in the optimization, making the speed approximately constant with value \(\hat{\lambda}\). This, in turn, makes the covered arc length function approximately linear (\(\apply{\lambda}{t} \approx \hat{\lambda}t\)), facilitating the simple statement of the optimization goals for the specification items and the fairness measure. It also makes the segment for each specification item fixed. Additionally, it causes the total arc length of the curve to be approximately equal to \(\hat{\lambda}\), as requested by the specification.

@@

				The error term given above is not well-suited for smooth optimization because it incorporates finding a maximum. Since the optimization problem will allow the original error term to be brought close to zero, it can be approximated using the following error term:
				\begin{equation*}
					\hat{\lambda}^{-2}\integral{0}{1}{\xp{\apply{\sigma}{t} - \hat{\lambda}}^2}{t}
				\end{equation*}
				As this term is intended to be used as a minimization goal, the factor \(\hat{\lambda}^{-2}\) was added, which makes it scale invariant. This enables the error term to be statically weighted against other terms in the objective function without this weighting being shifted due to curve scaling.

@@

				As mentioned in section \ref{section:description_language}, we are not concerned with properties beyond the second derivative of the curve. Thus, only \(\contg{2}\) continuity will be required. While the polynomial curves used as segments are smooth (\(\contp{\infty}\) continuity), the resulting spline in general is not. The segments have to be pieced together in such a way as to ensure a certain degree of smoothness. Thus, to ensure \(\contg{2}\) continuity for the whole curve, it is sufficient to ensure \(\contg{2}\) continuity at the segment connection points. To achieve this, the following error terms are added for each connection point (\(0 \leq i \leq m - 1\)):
				\begin{equation*}
					\begin{gathered}
						\apply{\phi_i}{1} - \apply{\phi_{i + 1}}{0}\\
						\apply{\phi'_i}{1} - \apply{\phi'_{i + 1}}{0}\\
						\apply{\phi''_i}{1} - \apply{\phi''_{i + 1}}{0}
					\end{gathered}
				\end{equation*}
				These error terms being zero implies \(\contp{2}\) continuity. However, since the constant speed requirement introduced in the previous paragraph collapses \(\contp{2}\) continuity and \(\contg{2}\) continuity, this is equivalent to the original requirement of ensuring only \(\contg{2}\) continuity, yet better suited for optimization.

				It is not acceptable for the curve segments to not touch, have sharp corners or curvature discontinuities, so the error terms above are added as zero-error constraints.

@@

				Having gathered all the necessary error terms, they are assembled in a nonlinear optimization problem. Terms with zero-error constraints, as well as terms with upper bounds on the error are added as component functions of the constraint function \(g\) together with appropriate bound components in \(g_l\) and \(g_u\).

				Minimization goals, on the other hand, are assembled in a weighted sum which defines the objective function \(f\). Since the error terms were designed to be scale invariant, this weighting can be done statically. The exact values of the weights are to be determined empirically.
